Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-16 ind_files/05_unification_discussion.tex
Minor rephrasing
Add alternative representation of stochastic integral terms

We can also calculate how the statistical mean value of any type-level quantity (see section \ref{sec_stat_measures}) changes over time. Let $f$ be any type level quantity, with value $f_i(t)$ for the $i\textsuperscript{th}$ type. I allow the value $f_i$ to possibly vary over time (for example, due to plasticity or a changing environment). By multiplying both sides of
equation \eqref{nD_stochastic_RM} by $f_i$ and summing over all $i$ (The same steps as going from \eqref{nD_replicator_mutator} to \eqref{nD_Price_time_dependent}), we see that the statistical mean $\overline{f}$ of the quantity in the population varies as:
\begin{equation}
\label{nD_stochastic_Price}
%The \vphantom is to vertically align the underbraces
\resizebox{\textwidth}{!}{$\displaystyle d\overline{f} = {\underbrace{\vphantom{ \frac{1}{KN_K(t)} }\textrm{Cov}(w,f)dt}_{\substack{\text{Classical} \\ \text{selection}}}} - {\underbrace{\frac{1}{KN_K(t)}\textrm{Cov}(\tau,f)dt}_{\substack{\text{Noise-induced} \\ \text{selection}}}} +  {\underbrace{\vphantom{ \frac{1}{KN_K(t)} }\overline{\left(\frac{\partial f}{\partial t}\right)}dt}_{\substack{\text{Ecological} \\ \text{feedbacks}}}}
+ \underbrace{\vphantom{ \frac{1}{KN_K(t)} }M_{\overline{f}}(\mathbf{p},N_K)dt}_{\substack{\text{Mutational} \\ \text{effects}}}
+ \underbrace{\frac{1}{\sqrt{K}N_{K}(t)}dW_{\overline{f}}}_{\substack{\text{Stochastic} \\ \text{fluctuations}}}$}
\end{equation}
where all covariances are understood in the statistical sense (see section \ref{sec_stat_measures}). Here,
\begin{equation}
	M_{\overline{f}}(\mathbf{p},N_K) \coloneqq \mu\left(1-\frac{1}{KN_K(t)}\right)\left(\sum\limits_{i=1}^{m}f_iQ_i(\mathbf{p}) - \overline{f}\sum\limits_{i=1}^{m}Q_i(\mathbf{p})\right)
\end{equation}
is a term capturing mutational effects/transmission fidelity that vanishes in the low mutation rate ($\mu \to 0$) limit, and 
\begin{equation}
	dW_{\overline{f}} \coloneqq \sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)\sqrt{A_i^+}dW_{t}^{(i)}
\end{equation}
is a stochastic integral term describing un-directed stochastic fluctuations and vanishes upon taking expectation values. Note that the $1/\sqrt{K}$ factor outside this term in equation \eqref{nD_stochastic_Price} means that these stochastic fluctuations also disappear in the infinite population ($K \to \infty$) limit.

@@

\subsection{The fundamental equation for the variance of a type-level quantity}\label{sec_fun_theorems_var}
Equation \eqref{nD_stochastic_Price} is a general equation for the mean value of an arbitrary type level quantity $f$ in the population. In many real-life situations, especially those pertaining to finite populations, we are interested in not just the mean, but also the variance of a type-level quantity. In Appendix \ref{App_stoch_var_eqns}, I show that the statistical variance of any type level quantity $f$ obeys
\begin{equation}
\label{nD_stochastic_Price_variance}
\begin{aligned}
d\sigma^2_{f} &= \textrm{Cov}\left(w,(f - \overline{f})^2\right)dt - \frac{1}{KN_K}\left[ \ \overline{\tau}\sigma^2_{f} +  2\textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \ \right]dt\\[12pt]
& + 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + \frac{1}{\sqrt{K}N_{K}(t)}dW_{\sigma^2_{f}}
\end{aligned}
\end{equation}
where
\begin{equation}
\label{variance_price_mutation_term}
M_{\sigma^2_f}(\mathbf{p},N_K) \coloneqq \mu\left[\left(1 - \frac{2}{KN_K}\right)\left(\sum\limits_{i=1}^{m}(f_i - \overline{f})^2Q_i(\mathbf{p})\right) + \sigma^2_f\left(1 - \frac{1}{KN_K}\right)\sum\limits_{i=1}^{m}Q_i(\mathbf{p})\right]
\end{equation}
is a mutational term that vanishes in the $\mu \to 0$ limit and
\begin{equation}
\label{variance_price_diffusion_term}
dW_{\sigma^2_f} \coloneqq \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2\sqrt{A_i^+}dW_{t}^{(i)}
\end{equation}
is a stochastic integral term measuring the (non-directional) effect of stochastic fluctuations that vanishes upon taking an expectation over the probability space. In the case of one-dimensional quantitative traits, an infinite-dimensional version of \eqref{nD_stochastic_Price_variance} has recently been rigorously derived~\citep{week_white_2021} using measure-theoretic tools under certain additional assumptions (See equation (21c) in~\cite{week_white_2021}). Taking expectations over the probability space in \eqref{nD_stochastic_Price_variance} and substituting mutation as acting via a Gaussian kernel also recovers equations previously derived~\citep{debarre_evolutionary_2016} in the context of evolutionary branching in finite populations as a special case. An infinite population ($K \to \infty$) version of equation \eqref{nD_stochastic_Price_variance} also appears in~\cite{lion_theoretical_2018}.

>>>>

We can also calculate how the statistical mean value of any type-level quantity (see section \ref{sec_stat_measures}) changes over time. Let $f$ be any type level quantity, with value $f_i(t) \in \mathbb{R}$ for the $i\textsuperscript{th}$ type. I allow the value $f_i$ to possibly vary over time (for example, due to plasticity or a changing environment). By multiplying both sides of
equation \eqref{nD_stochastic_RM} by $f_i$ and summing over all $i$ (The same steps as going from \eqref{nD_replicator_mutator} to \eqref{nD_Price_time_dependent}), we see that the statistical mean $\overline{f}$ of the quantity in the population varies as:
\begin{equation}
\label{nD_stochastic_Price}
%The \vphantom is to vertically align the underbraces
\resizebox{\textwidth}{!}{$\displaystyle d\overline{f} = {\underbrace{\vphantom{ \frac{1}{KN_K(t)} }\textrm{Cov}(w,f)dt}_{\substack{\text{Classical} \\ \text{selection}}}} - {\underbrace{\frac{1}{KN_K(t)}\textrm{Cov}(\tau,f)dt}_{\substack{\text{Noise-induced} \\ \text{selection}}}} +  {\underbrace{\vphantom{ \frac{1}{KN_K(t)} }\overline{\left(\frac{\partial f}{\partial t}\right)}dt}_{\substack{\text{Ecological} \\ \text{feedbacks}}}}
+ \underbrace{\vphantom{ \frac{1}{KN_K(t)} }M_{\overline{f}}(\mathbf{p},N_K)dt}_{\substack{\text{Mutational} \\ \text{effects}}}
+ \underbrace{\frac{1}{\sqrt{K}N_{K}(t)}dW_{\overline{f}}}_{\substack{\text{Stochastic} \\ \text{fluctuations}}}$}
\end{equation}
where all covariances are understood in the statistical sense (see section \ref{sec_stat_measures}). Here,
\begin{equation}
	M_{\overline{f}}(\mathbf{p},N_K) \coloneqq \mu\left(1-\frac{1}{KN_K(t)}\right)\left(\sum\limits_{i=1}^{m}f_iQ_i(\mathbf{p}) - \overline{f}\sum\limits_{i=1}^{m}Q_i(\mathbf{p})\right)
\end{equation}
is a term capturing mutational effects/transmission fidelity that vanishes in the low mutation rate ($\mu \to 0$) limit, and 
\begin{equation}
	dW_{\overline{f}} \coloneqq \sum\limits_{i=1}^{m}\left(f_i-\overline{f}\right)\sqrt{A_i^+}dW_{t}^{(i)}
\end{equation}
is a stochastic integral term describing un-directed stochastic fluctuations and vanishes upon taking expectation values (Also see equation \eqref{nD_stoch_Price_noise_term_alt_representation} in Appendix \ref{App_alternative_representation} for a more elegant representation of this term as an integral against a single Wiener process). Note that the $1/\sqrt{K}$ factor outside this term in equation \eqref{nD_stochastic_Price} means that these stochastic fluctuations also disappear in the infinite population ($K \to \infty$) limit.

@@

\subsection{The fundamental equation for the variance of a type-level quantity}\label{sec_fun_theorems_var}
Equation \eqref{nD_stochastic_Price} is a general equation for the mean value of an arbitrary type level quantity $f$ in the population. In many real-life situations, especially those pertaining to finite populations, we are interested in not just the mean, but also the variance of a type-level quantity. In Appendix \ref{App_stoch_var_eqns}, I show that the statistical variance of any type level quantity $f$ obeys
\begin{equation}
\label{nD_stochastic_Price_variance}
\begin{aligned}
d\sigma^2_{f} &= \textrm{Cov}\left(w,(f - \overline{f})^2\right)dt - \frac{1}{KN_K}\left[ \ \overline{\tau}\sigma^2_{f} +  2\textrm{Cov}\left(\tau,(f - \overline{f})^2\right) \ \right]dt\\[12pt]
& + 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + \frac{1}{\sqrt{K}N_{K}(t)}dW_{\sigma^2_{f}}
\end{aligned}
\end{equation}
where
\begin{equation}
\label{variance_price_mutation_term}
M_{\sigma^2_f}(\mathbf{p},N_K) \coloneqq \mu\left[\left(1 - \frac{2}{KN_K}\right)\left(\sum\limits_{i=1}^{m}(f_i - \overline{f})^2Q_i(\mathbf{p})\right) + \sigma^2_f\left(1 - \frac{1}{KN_K}\right)\sum\limits_{i=1}^{m}Q_i(\mathbf{p})\right]
\end{equation}
is a mutational term that vanishes in the $\mu \to 0$ limit and
\begin{equation}
\label{variance_price_diffusion_term}
dW_{\sigma^2_f} \coloneqq \sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2\sqrt{A_i^+}dW_{t}^{(i)}
\end{equation}
is a stochastic integral term measuring the (non-directional) effect of stochastic fluctuations that vanishes upon taking an expectation over the probability space (Also see equation \eqref{nD_stoch_variance_Price_noise_term_alt_representation} in Appendix \ref{App_alternative_representation} for a more elegant representation of this term as an integral against a single Wiener process). In the case of one-dimensional quantitative traits, an infinite-dimensional version of \eqref{nD_stochastic_Price_variance} has recently been rigorously derived~\citep{week_white_2021} using measure-theoretic tools under certain additional assumptions (See equation (21c) in~\cite{week_white_2021}). Taking expectations over the probability space in \eqref{nD_stochastic_Price_variance} and substituting mutation as acting via a Gaussian kernel also recovers equations previously derived~\citep{debarre_evolutionary_2016} in the context of evolutionary branching in finite populations as a special case. An infinite population ($K \to \infty$) version of equation \eqref{nD_stochastic_Price_variance} also appears in~\cite{lion_theoretical_2018}.
